---
title: "Bellabeat Cleaning Report"
author: "Alex T"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages
```{r}
#Load necessary packages
library(tidyverse)
library(janitor)
library(skimr)
library(dplyr)
library(readr)
```

### Importing data
Cleaning began with changing the csv file names in the data set to be more
uniform. Words were broken up by underscores and words were made all lowercase.
This step was done via Google Docs before the files were imported into RStudio.

The FitBit tracker data set included 18 csv files. After skimming the data in 
Google Sheets, I decided on the following data sets: 
```{r}
#Create data frames from csv files with clean names
daily_activity_df <- clean_names(read.csv('daily_activity_merged.csv'))
daily_calories_df <- clean_names(read.csv('daily_calories_merged.csv'))
hourly_intensity_df <- clean_names(read.csv('hourly_intensities_merged.csv'))
daily_intensity_df <- clean_names(read.csv('daily_intensities_merged.csv'))
daily_steps_df <- clean_names(read.csv('daily_steps_merged.csv'))
heartrate_df <- clean_names(read.csv('heartrate_seconds_merged.csv'))
sleep_df <- clean_names(read.csv('sleep_day_merged.csv'))
weight_df <- clean_names(read.csv('weight_log_info_merged.csv'))
```

The imported csv files were then added to a list for easier management.
```{r}
#Create list to hold all data frames
data_list = list('daily_activity '= daily_activity_df,
                 'daily_calories' = daily_calories_df,
                 'hourly_intensity' = hourly_intensity_df,
                 'daily_intensity' = daily_intensity_df,
                 'daily_steps' = daily_steps_df,
                 'heartrate' = heartrate_df,
                 'sleep' = sleep_df,
                 'weight' = weight_df)
```

Utilized the 'head()' function on all the data sets to to get a quick peek at
everything before proceeding. 
```{r}
lapply(data_list, function(x) {
  head(x)
})
```

The information on the FitBit data set said that it included data from 30 users 
so I decided to double check that fact.
```{r}
#Get the distinct number of users in each data set
lapply(data_list, function(x) {
  print(n_distinct(x$id))
})

```

Upon running the above function, I discovered that there was actually data for
33 users in 5 of the data sets(daily activity, daily calories, hourly intensity,
daily intensity, and daily steps).There was data for 14 users in the heart rate
data set, 24 in the sleep data set, and 8 in the weight data set. There weren't
enough users in the heart rate or weight data sets to make any decisions with
confidence so they were excluded from the analysis going forward.
```{r}
#Remove list elements
data_list <- within(data_list, rm('heartrate'))
data_list <- within(data_list, rm('weight'))
length(data_list)
```

The list of data frames was then looped through in order to find incorrect data
values.
```{r}
#Check all rows for data validity
lapply(data_list, function(x) {
  lapply(x, function(y) {
    #Check for empty or NULL values
    if (is.null(y) | is_empty(y)) {
      print("NULL or empty values")
    }
    #Run checks on columns to ensure values are the correct data type
     else if (typeof(y) == "double" | typeof(y) == "integer") {
      if (is.numeric(y)) {
        print("All numerical data types are correct")
      } else {
        print ("Incorrect numerical data types")
      }
    }
    else if (typeof(y) == "character") {
      if (is.character(y)) {
        print("All character data types are correct")
      } else {
        print ("Incorrect character data types")
      }
    }
  })
})
```

Now to check for duplicated values and to remove them if any are found
```{r}
#Check for duplicate values
sapply(data_list, function(x) sum(duplicated(x)))

#Remove duplicate values from sleep data set
data_list$sleep <- data_list$sleep[!duplicated(data_list$sleep), ]

sapply(data_list, function(x) sum(duplicated(x)))
```

I noticed that the hourly intensity and sleep data sets included the time along
with the date in the same column, so I split them into their own separate
columns.

The time values in the old sleep column
were all 12 am so there was no need to include that in a new column.
```{r}
#split date and time into their own respective columns
data_list$hourly_intensity$activity_date <- (format(as.Date(data_list$hourly_intensity$activity_hour, '%m/%d/%Y'), format ="%m/%d/%Y"))
data_list$hourly_intensity$activity_time <- ((format(as.POSIXct(data_list$hourly_intensity$activity_hour, '%m/%d/%Y %I:%M:%S %p', tz =''), format = "%I:%M %p")))
data_list$sleep$sleep_date <-(format(as.Date(data_list$sleep$sleep_day, '%m/%d/%Y'), format ="%m/%d/%Y"))
```

The old data columns can now be dropped.
```{r}
#Remove old date columns
data_list$hourly_intensity <- subset(data_list$hourly_intensity, select = -activity_hour)
data_list$sleep <- subset(data_list$sleep, select = -sleep_day)

head(data_list$hourly_intensity)
head(data_list$sleep)
```


Adjusted the column names in the other data sets to 'date' instead of 'day'.
```{r}
#replace columns with suffix 'day' with 'date'
colnames(data_list$daily_calories) <- gsub(colnames(data_list$daily_calories), pattern = '_day', replacement = '_date')
colnames(data_list$daily_intensity) <- gsub(colnames(data_list$daily_intensity), pattern = '_day', replacement = '_date')
colnames(data_list$daily_steps) <- gsub(colnames(data_list$daily_steps), pattern = '_day', replacement = '_date')
```

Adjust the date format in the other data sets to also have a two digit month
```{r}
#Format the other dates
data_list$`daily_activity `$activity_date <- format(mdy(data_list$`daily_activity `$activity_date), format = '%m/%d/%Y')
data_list$`daily_calories`$activity_date <- format(mdy(data_list$`daily_calories`$activity_date), format = '%m/%d/%Y')
data_list$`daily_intensity`$activity_date <- format(mdy(data_list$`daily_intensity`$activity_date), format = '%m/%d/%Y')
data_list$`daily_steps`$activity_date <- format(mdy(data_list$`daily_steps`$activity_date), format = '%m/%d/%Y')

sapply(data_list, head)
```


The minimum for the calories columns are 0.This should be impossible according
to the Basal Metabolic Rate(BMR),which is basically the number of calories
you burn just by existing. The zeroes will be converted to NA.
```{r}
summary(data_list$daily_calories$calories)

#Convert the 0 values into NA
data_list$daily_calories[data_list$daily_calories == 0] <- NA
data_list$`daily_activity `$calories[data_list$daily_activity$calories == 0] <- NA

summary(data_list$daily_calories$calories)
```

```{r}
saveRDS(data_list, 'data_list.rds')
```