---
title: "Bellabeat Case Study"
author: "Alex T"
date: "2023-04-12"
output: html_document
---
```{r echo=FALSE, message=FALSE}
#Load necessary packages
library(tidyverse)
library(janitor)
library(skimr)
library(dplyr)
library(readr)
```


## About the company
Bellabeat is a high-tech company that manufactures health-focused
smart products.Bellabeat has grown rapidly since its founding in 2013 and has
centered itself as a tech-driven wellness company for women.


## Business task
Identify possible growth opportunities for Bellabeat smart devices and marketing
strategies based on insights from other health-focused smart devices.


## The data
The data set for this case study was the FitBit Fitness Tracker Data.
The data set was provided by user Mobius through the website Kaggle and was
generated originally by respondents to a distributed survey via Amazon
Mechanical Turk. The survey ran for 61 days starting March,12th 2016 and ending
on May, 12th 2016. Thirty eligible FitBit users consented to the submission of
personal tracker data, including minute-level output for physical activity,
heart rate, and sleep monitoring.

## Diving into the data
Summary of the data sets:
```{r echo=FALSE}
con <- gzfile("data_list.rds")
data_list <- readRDS(con)

#Get summary of data
lapply(data_list, function(x) {
  select(x, (!ends_with('id') & !ends_with('date') & !ends_with('time'))) %>% 
    summary
})

close(con)

#Remove scientific notation
options(scipen=999)

#change name of sleep date column for merging purposes
colnames(data_list$sleep)[5] <- 'activity_date'

#merge activity and sleep data frames
merged_activity_sleep <- merge(data_list$`daily_activity `,
                               data_list$sleep, by=c ('id', 'activity_date'))
```
### Key details from summary

* Users sleep once on average for 419.5 minutes(7 hours).  
  + 7 hours is the recommended hours of sleep for adults aged 12 - 60.  
&nbsp;    

* The average steps per day is 7638.
  + According to the CDC, 8000 steps per day is associated with a 51% lower risk
    for all-cause mortality, or death from all causes.
  + 12,000 steps per day is associated with a 65% lower risk.
  + The number of steps has a strong association with morality, not the
    intensity.  
&nbsp;   
* A majority of the users are lightly active when not sedentary.
&nbsp;   
* Users on average are sedentary for 991 minutes(16.5 hours). Could imply that
  the individuals chosen for this data set work office/desk jobs.
    
### Looking at the data 
```{r, echo=FALSE}
#intensity and time
intensity_time <- data_list$hourly_intensity
#Change the display order for the x axis
intensity_time$activity_time <- factor(intensity_time$activity_time,
                                       c("01:00 AM", "02:00 AM",
                                         "03:00 AM", "04:00 AM",
                                         "05:00 AM","06:00 AM",
                                         "07:00 AM", "08:00 AM",
                                         "09:00 AM", "10:00 AM",
                                         "11:00 AM", "12:00 PM",
                                         "01:00 PM", "02:00 PM",
                                         "03:00 PM", "04:00 PM",
                                         "05:00 PM","06:00 PM",
                                         "07:00 PM", "08:00 PM",
                                         "09:00 PM", "10:00 PM",
                                         "11:00 PM", "12:00 AM"))
intensity_time <- intensity_time%>% 
  group_by(activity_time) %>% 
  summarise(intensity = sum(total_intensity))

ggplot(intensity_time, aes(x=activity_time, y=intensity, fill='#FE8F77')) +
  geom_col(show.legend = FALSE) +
  labs(title = 'Activity by the hour', x = 'Hour', y = 'Total intensity') +
  theme(axis.text.x = element_text(angle = 90))
```

Looking at the chart we can see that:  

* People are the most active during the hours of 5-7 PM. It would seem that most
  seem to begin their workout after getting off work.  
* People favor early morning workouts over late night.
  
```{r echo=FALSE, message=FALSE}

#change name of sleep date column for merging purposes
colnames(data_list$sleep)[5] <- 'activity_date'

#merge activity and sleep data frames
merged_activity_sleep <- merge(data_list$`daily_activity `,
                               data_list$sleep, by=c ('id', 'activity_date'))
#Calories and steps
steps_calories <- merged_activity_sleep %>% 
  group_by(total_steps) %>% 
  summarise(calories_burned = sum(calories))

ggplot(steps_calories, aes(x=total_steps, y=calories_burned)) +
  geom_smooth() +
  geom_point(stat = 'identity') +
  labs(title = 'Steps and calories correlation',
       x = 'Total steps', y = 'Calories')
```

* We can see that there is a positive correlation between calories burned
  and steps taken. The more active you are, the more calories you burn.  
* According to the CDC, intensity doesn't really matter in this case so we
  can let app users know they are allowed to pace themselves with their steps
  throughout the day. 
  
  
  
```{r echo=FALSE}
#Daily steps
daily_steps <- data_list$daily_steps
daily_steps$weekday <- wday(mdy(daily_steps$activity_date), label = TRUE)
daily_steps$weekday <- factor(daily_steps$weekday, c("Mon", "Tue", "Wed", "Thu","Fri","Sat", "Sun"))

daily_steps <- daily_steps%>% 
  group_by(weekday) %>% 
  summarise(total = mean(step_total))

ggplot(daily_steps, aes(x=weekday, y=total,  fill='#FE8F77')) +
  geom_col(show.legend = FALSE) + geom_hline(yintercept = 8000) +
  labs(title = "Average steps per day", x= 'Weekday', y = 'Steps')
```
  

Looking at the chart tells us:  

* On average, the CDC recommended step goal is only reached on Tuesday & Saturday
    + While 8000 steps are good, 10,000 or 12,000 steps daily would be ideal per
    the CDC.
* The lowest average steps on Sunday.
    + Could be because people are mainly relaxing and getting ready for the work
    week.
    
## Recommendations

### Recommendations for future analysis
* A more detailed and diverse data set would be extremely beneficial to gathering
more accurate insights. Future data should include the user's sex, height, and 
age.it could also be helpful to include the type of work they do categorized by
activity level: sedentary, moderate etc.

* The data should also feature a larger number of participants with varying
characteristics such as age and sex.

* Having app users answer various survey questions after consistently using the
  app for at least 1 month would be very helpful with gathering insight for
  future analysis.
  + Survey questions could be about the app itself or anything pertaining to
    exercise.


### Recommendations for Bellabeat app
* The app should give users the ability to turn on health notifications/tips
  that are separate from the standard in-app notifications. Notifications could
  be from the CDC or any reputable health organization.
  
  + Notifications could be sent daily, weekly, or even dependent on particular
    situations(E.g: step counter approaching the recommended 8000,
    sleeping less than the recommended hours). The goal here is to inform users
    on why it may be important to hit certain numbers.
    
  + Tips and techniques tailored to an individual's stats should also be sent
    out periodically.
    
  + The membership program may have to be reworked in order to prevent the free
    insights from compromising on the paid insights.Analysis on the
    membership program would be needed to discuss further.
    
* Using the app consistently should allow users to receive discount on other
  Bellabeat products.
  
### General recommendations
* Having ads play during the busiest workout times(5PM - 7PM) could possibly be
  beneficial to getting more people to use the Bellabeat app/products if the
  busy trend is reflective of other people that workout.
    + Radio ads for those getting off work or on the way to the gym.
      (Spotify included)
    + Television ads for gyms that have accessible TVs.

## Conclusion
While the current data will suffice at the moment, more data will be needed in
order to give more detailed insight and to make bigger business decisions with
confidence. A survey involving current Bellabeat customers would be the best way
to springboard any further phases of analysis.


## Appendix
* [CDC Recommended sleep](https://www.cdc.gov/sleep/about_sleep/how_much_sleep.html)
* [CDC Steps per day](https://www.cdc.gov/media/releases/2020/p0324-daily-step-count.html#:~:text=Taking%2012%2C000%20steps%20per%20day,compared%20with%20taking%204%2C000%20steps)